---
title: "Hierarchical models and sequential model fits"
author: "Leontine"
format: html
editor: visual
---

# Overview

In this document we explain

1.  how we set up a hierarchical model for a parameter; this code set up was introduced in <https://github.com/AlkemaLab/BayesTransitionModels>.

2.  how we fix higher level parameters for that set up when doing sequential runs.

# Set up

Load relevant packages, including our own

```{r}
#| include: false
library(here)
library(tidyverse)
library(ggplot2)
library(haven)
library(stringr)
library(cmdstanr)
library(truncnorm) # for inits
library(bayesplot) # diagnostic plots
devtools::load_all(here::here())
```

Read in data and preprocess as usual

```{r}
#| include: false
data_folder <- "data_raw"
dat <- read_csv(here::here(data_folder, "coverage_data.csv"))
dat_subnat <- read_csv(here::here(data_folder, "coverage_data_subnat.csv"))

```

Let's use columns `cluster` and `subcluster` and `iso`:

```{r}
hierarchical_level <- c("intercept",  "subcluster", "iso")
dat %>% select(iso, cluster, subcluster)
```

Summary of clusters and subclusters, after selecting one row per country

```{r}
dat %>%
  group_by(iso) %>%
  filter(year == max(year)) %>%
  select(country, iso, subcluster, cluster) %>%
  group_by(cluster) %>%
  summarize(nsubclusters = length(unique(subcluster)), ncountries = n())
```

# Model fitting

## Model set up

Let's consider the following model (we won't bother with logit-transforms):

-   parameter to estimate = $\mu_c$ = country-specific coverage (during entire period)

-   data model: $y_i \sim N(\mu_{c[i]}, \sigma_y^2)$

-   hierarchical model for $\mu_c$:

    -   $\mu_c \sim N(\mu_{s[c]}^{(subregion)}, \sigma_{\mu, across-country}^2)$ ,

    -   $\mu_s^{(subregion)} \sim N(\mu_{r[s]}^{(region)}, \sigma_{\mu, across-subregions}^2)$,

    -   $\mu_r^{(region)} \sim N(\mu^{(global)}, \sigma_{\mu, across-regions}^2)$

Note that an equivalent specification of the hierarchical model (the one we'll use in the stan model) is as follows:

-   $\mu_c =  \mu^{(global)} + \eta_{r[s[c]]]}^{(region)} + \eta_{s[c]}^{(subregion)} + \eta_c$ with

-   $\eta_c \sim N(0, \sigma_{\mu, across-country}^2)$ ,

-   $\eta_s^{(subregion)} \sim N(0, \sigma_{\mu, across-subregions}^2)$,

-   $\eta_r^{(region)} \sim N(0, \sigma_{\mu, across-regions}^2)$

In addition, in the stan model, we will use that we can write $\eta = \sigma\cdot \gamma$ where $\gamma \sim N(0,1)$. In addition, with the same set up, if $\mu^{(global)}$ has prior

$$\mu^{(global)} \sim N(0, s_{\mu, global}^2),$$then we can write $\mu^{(global)} = s_{\mu, global}\cdot \gamma^{(global)}$, with $s$ being the fixed prior standard deviation (as opposed to a SD that's estimated). All that combined, we get

-   $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{\mu, region}\cdot \gamma_{r[s[c]]]}^{(region)} + \sigma_{\mu, subregion}\cdot\gamma_{s[c]}^{(subregion)} + \sigma_{\mu, country}\cdot\gamma_{\mu, c}$ with

-   $\gamma \sim N(0,1)$ for each gamma (one global, $R$ region, $S$ subregions, $C$ countries).

The set up and notation in stan/R, in terms of parameters, using parameter mu as an example, is currently as follows:

-   mu_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.

-   mu_sigma referred to the vector of stacked sigmas, excluding $s_{\mu, global}$ (in the original package, it was called mu_sigma_raw).

-   mu_star refers to the vector with (mu global and ALL $\eta$s). we get mu_stars from mu_raw and the vector ($s_{\mu, global}$ , mu_sigma), using the stan `scale_blocks` function.

-   mu refers to the final $\mu$s at the country level. Matrix multiplication is used to obtain the mu country vector from the mu_stars. The full matrix `mu_model_matrix` is a sparse matrix, with 1s and 0s (columns refer to $\mu_{global}$, the regional etas, the subregional etas, the country etas). The matrix multiplication is implemented using the stan function csr_matrix_times_vector, using a decomposition of `mu_model_matrix` into different parts U, V, W.

-   Because of the sequential fitting, for mu_raw and mu_sigma, we also had to distinguish between parameters that were fixed vs those that were estimated. We do so with the \_fixed and \_estimate additions, i.e. the vector of mu_raws is produced from mu_raw_fixed and mu_raw_estimate.

## Model fitting

global model fit

```{r}
fit_global <- fit_model_localhierarchy(runstep = "global_national",
                              hierarchical_level     =  hierarchical_level,
                              survey_df = dat,
                              chains = 4)

```

create summary object

```{r}
fit_global$post_summ <- get_posterior_summaries(fit_global)
fit_global$post_summ

```

do local fit... here we fix things but do get results for all countries

```{r}
fit_local <- fit_model_localhierarchy(runstep = "local_national",
                                  global_fit = fit_global,
                                  survey_df = dat,
                                  chains = 4)
```

#### Produce and check outputs

get summaries, takes a little while. this creates, per fit, a list with each entry being one level, and the results (mu estimates) for that level

```{r}
res_global <- posterior_summary_hierparam(fit = fit_global, parname = "mu")
res_local <- posterior_summary_hierparam(fit = fit_local, parname = "mu")

```

Plots that show outputs

```{r}
plot_posterior_summaries(res = res_global, modelname1 = "global")
```

how values have been fixed in the local runs

```{r}
plot_posterior_summaries(res = res_local, 
                         res2 = res_global,
                         modelname2 = "global", modelname1 = "local") 

```

selecting specific levels

```{r}
plot_posterior_summaries(res = res_local, 
                         res2 = res_global,
                         modelname2 = "global", modelname1 = "local",
                        hierarchy_select = "subcluster")



```

Level and selected areas

```{r}
plot_posterior_summaries(res = res_local, 
                         res2 = res_global,
                         modelname2 = "global", modelname1 = "local",
                        hierarchy_select = "iso",
                        areas_select = res_local$iso$name[1:100])

```

# Summary object

Going through what's happening to create the summary object

Set up hierarchical structures, folow the hierdata.qmd

```{r}
area <- "iso"
```

Create district index for matching district and district index

```{r}
geo_unit_index <- get_geo_unit_index_data(dat,
                                          hierarchical_levels = c(hierarchical_level), 
                                          area = area)
geo_unit_index
```

```{r}
hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
names(hier_data)
hier_data


```

This is where global fit is used to fix values

```{r}
hier_stan_data <- hierarchical_param_stan_data(
  global_fit = fit_global,
  param_name = "mu",
  param_data = hier_data,
  hierarchical_terms_fixed = hierarchical_level[1:2], 
  hierarchical_sigmas_fixed = hierarchical_level[1:3]
)
names(hier_stan_data)
```

# Prior-posts 

sigmas

```{r}
plot_prior_post_sigmas(fit = fit_globalsubnat, parname = "mu")

```

create plots for mu_raws

```{r}
plots <- plot_mu_raw(fit = fit_global, parname = "mu")
# a list with plots
```

summary plots gives summaries, 30 at a time

```{r}
plots[["summary_plots"]][[1]]

```

specific prior-post plots for one parameter are in plots\[\["plots_allmuraw"\]\]

```{r}
plots[["plots_allmuraw"]][[1]]
```

Simple trace plots etc

```{r}

fit <- fit_global
 parname_list <- list( c("mu_raw_estimate[1]",
                          "mu_sigma_estimate"))
#    plot_dirandname <- file.path(fit$output_dir, paste0(runname, "diagnostics.pdf"))
#    pdf(plot_dirandname, width = 11, height = 11)
    for (i in 1:length(parname_list)){
      parname <- parname_list[[i]]
      samp <- fit$samples$draws(parname )
      #mcmc_trace(samp)
      p <- mcmc_dens_overlay(samp)
      # for more than 1 parameter
      #mcmc_pairs(samp)
      print(p)
    }
#    dev.off()
```
