---
title: "hierarchical data"
author: "leontine"
format: html
editor: visual
---

# 

# Demonstration of hierarchical data with toy data set.

Load relevant packages

```{r}
#| include: false
library(here) 
library(tidyverse) 
library(ggplot2) 
library(haven) 
library(stringr)
library(cmdstanr) 
library(bayesplot) # diagnostic plots 
devtools::load_all(here::here())
```

Toy data set

```{r}
data <- tibble(subcluster = c("A", "A", "B", "B"), iso = c("iso1", "iso2", "iso3", "iso3"))
data
```

We need to define the hierarchical levels to use, and the level of interest. Here we set level of interest to be equal to lowest level (iso), this is currently required (and the default setting in all runs in fit_model_localhierarchy).

(This does not need to be a requirement going forward but testing so far is based on it).

```{r}
hierarchical_level <- c("intercept",  "subcluster", "iso")
area <- "iso"
```

Create district index for matching district and district index

```{r}
geo_unit_index <- get_geo_unit_index_data(data,
                                          hierarchical_levels = c(hierarchical_level), 
                                          area = area)
geo_unit_index
```

Set up hierarchical structures

```{r}
hier_data <- hierarchical_data(geo_unit_index, hierarchical_level)
names(hier_data)
hier_data


```

for use in stan

```{r}
hier_stan_data <- hierarchical_param_stan_data(
  global_fit = NULL,
  param_name = "mu",
  param_data = hier_data #,
  # used only when fixing things 
#  hierarchical_terms_fixed = hierarchical_level_terms_fixed,
#  hierarchical_sigmas_fixed = hierarchical_level_sigmas_fixed
)
names(hier_stan_data)
```

```{r}
hier_stan_data$mu_raw_n_terms # number of etas + global mu
length(unique(geo_unit_index$subcluster))
length(unique(geo_unit_index$iso))
  

```

```{r}
hier_stan_data$mu_n_sigma # update: no longer includes prior sd for intercept 
```

Let's illustrate how these objects are used in R and in the stan model.

For that, make up some values for the parameters based on

```         
"mu_n_terms"          "mu_n_sigma"
```

```{r}
#mu_raw refers to the vector with ALL $\gamma$s, including the global one.The ordering is global-allregional-allsubregional-country.
mu_raw <- seq(1, hier_stan_data$mu_raw_n_terms)
mu_raw
#mu_sigma referred to the vector of stacked sigmas, excluding $s_{\mu, global}$ (in the #original package, it was called mu_sigma_raw).
mu_sigma <- seq(1, hier_stan_data$mu_n_sigma)
sprior <- 0.5


```

try to expose stan functions using cmdstanr (not rstan)

<https://mc-stan.org/cmdstanr/reference/model-method-expose_functions.html>

```{r}
stan_file <- here("inst/stan/hierfunctions_seq_muscalar.stan")
mod <- cmdstan_model(stan_file, force_recompile = TRUE)  
mod$expose_functions(global = TRUE) # to add to global environment 

```

mu_star refers to the vector with (mu global and ALL $\eta$s). we get mu_stars from mu_raw \# and the vector ($s_{\mu, global}$ , mu_sigma), using the stan scale_blocks function.

vector scale_blocks(vector mu_raw, int mu_raw_n_terms, vector mu_sigmawpriorsd, int n_mu_sigmawpriorsd, array\[\] int re_start, array\[\] int re_end) {

```{r}
mu_star <- scale_blocks(mu_raw = mu_raw, 
                        mu_raw_n_terms = hier_stan_data$mu_raw_n_terms, 
                        mu_sigmawpriorsd = c(sprior, mu_sigma),
                        n_mu_sigmawpriorsd = hier_stan_data$mu_n_sigma +1, 
             re_start = hier_stan_data$mu_re_start,
             re_end = hier_stan_data$mu_re_end)
mu_star
```

mu refers to the final $\mu$s at the country level. Matrix multiplication is used to obtain the mu country vector from the mu_stars. The full matrix mu_model_matrix is a sparse matrix, with 1s and 0s (columns refer to $\mu_{global}$, the regional etas, the subregional etas, the country etas). The matrix multiplication is implemented using the stan function csr_matrix_times_vector, using a decomposition of mu_model_matrix into different parts U, V, W.

check if this is right

```{r}
mu <- hier_stan_data$mu_model_matrix%*%mu_star
mu
```

# 

```{r}
hier_data$model_matrix
```

# Notes on unsuccessful investigation into exposing stan build-in functions

If we want to use all stan functions, we need to expose (compile) the stan math functions as well, eg for for sparse matrices. see here for info

<https://mc-stan.org/docs/functions-reference/sparse_matrix_operations.html#conversion-functions>

<https://stan-dev.r-universe.dev/articles/StanHeaders/stanmath.html>

<https://discourse.mc-stan.org/t/expose-stan-functions-with-stan-math-library-function/15771/6>

help(stanFunction, package = "StanHeaders")

first time around, i got an error that I tried to fix

Error in sourceCpp(code = code, env = env, rebuild = rebuild, cacheDir = cacheDir, : Error 1 occurred building shared library.

But it turns out that it may not work anyway:

looks like csr_extract_w would be tricky... (was getting errors anyway with something along lines of StanHeaders::stanFunction("csr_extract_w"))

<https://discourse.mc-stan.org/t/function-to-return-number-of-non-zero-elements-in-a-matrix/26123/7>

so moving forward in R with (sparse) model_matrix\*vector

```{r}
#csr_extract_w()
```

```{r}
# get_onehierparam(oneparam_sigma = c(sprior, mu_sigma),  
#                  oneparam_raw_fixed = mu_raw[1], #numeric(0), 
#                  oneparam_raw_estimate = mu_raw[-1],
#                  oneparam_re_start = hier_stan_data$mu_re_start,
#                  oneparam_re_end = hier_stan_data$mu_re_end, 
#                  C = nrow(geo_unit_index), 
#                  oneparam_n_terms = hier_stan_data$mu_n_terms, 
#                  oneparam_n_terms_fixed = 1, #hier_stan_data$mu_n_terms_fixed, 
#                  oneparam_n_terms_estimate = hier_stan_data$mu_n_terms_estimate -1,
#       oneparam_model_matrix_w= csr_extract_w(hier_stan_data$mu_model_matrix), 
#       oneparam_model_matrix_v = hier_stan_data$mu_model_matrix_v, 
#       oneparam_model_matrix_u = hier_stan_data$mu_model_matrix_u
#           )
```
